Extended Data

### 2022-03-12 doc/manuscript/Dow_manuscript_Nature.Rmd
refining text, citations

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
## potential reviewers: Zani, Keenan

@@

<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at 3005 words and 1 modest + 2 large display items. Articles typically have 30-50 references (currently at 59)-->

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010-->;@menzel_european_2006;<!--@menzel_growing_1999-->;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

@@

<!--summary-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014-->;@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to high $T_{max}$ during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].

@@

Although stem diameter increments are routinely used to calculate $ANPP_{woody}$[@refs], stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016; @and_refs_therein].  
Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006-->;@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006-->;@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
*Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that $ANPP_{woody}$ is limited by sink factors, such that the expected increase photosynthate availability[@refs] (observed at Harvard Forest[@finzi_carbon_2020_but_verify]) does not enhance growth.*
*Indeed, such a phenomenon has been observed when GPP has been increased through lengthening of the growing season[@refs] or elevated CO~2~[@refs].*

<!--on the fate of extra C-->
Our results add to a growing body of evidence for a sink limitation of stem growth and $ANPP_{woody}$[@refs], wherein global change factors such as warming spring temperatures[@refs] or elevated CO~2~[@refs] enhance GPP, but woody growth is less responsive, if at all[@refs].
The fate of any additional C fixed during years with warm springs remains unresolved, but possible *destinations* -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, in a situation where GPP was enhanced (through experimental elevation of CO~2~) but $ANPP_{woody}$ was constrained *(presumably by sink factors)*, additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C loss through fall or winter respiration can offset gains from an earlier spring[@oishi_warmer_2018;@buermann_widespread_2018].
It is also possible that a buildup of NSCs that are not required for growth or other functions would feed back to reduce the rate of photosynthesis[@refs], such that sink limitation of $ANPP_{woody}$ reduces GPP relative to its potential in years with warm springs.
Either way, our results demonstrate that warm springs do not increase $ANPP_{woody}$ on an interannual time scale, contrasting with their tendency to augment _NEE_[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014].

@@

<!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season. 
Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig. `r tree_rings`, `r EDT_tree_ring_summary`).
This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress[@refs] or high temperatures[@anderson-teixeira_joint_2021] (which cause drought stress[@refs]) during the peak growing season.
<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@refs].

>>>>

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}

@@

<!--LENGTH: The typical length of an article with 3-4 modest display items (figures and tables) is 2000-2500 words (summary paragraph plus body text). We're currently at 2839 words and 1 modest + 2 large display items. Articles typically have 30-50 references (currently at 53)-->

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;-->@menzel_european_2006;<!--@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
However, model representation of C allocation to stem growth is overly simplistic and does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of $ANPP_{woody}$ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

@@

<!--summary-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase $ANPP_{woody}$, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;<!--@fu_variation_2014;-->@delpierre_wood_2016;@delpierre_temperate_2016].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to high $T_{max}$ during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022].

@@

Although stem diameter increments are routinely used to calculate $ANPP_{woody}$, stem volume growth does not directly translate into C content, but rather can be partially decoupled through differences in wood density or C content[@delpierre_wood_2016].  
Thus, it is theoretically possible that extra C is allocated to enhanced cell wall thickening, a process that lags behind stem expansion[@cuny_woody_2015], or to a higher ratio of high-density latewood to lower-density earlywood, resulting in formation of more C-dense wood in years with warm springs.
However, existing evidence indicates that warm springs have a neutral or negative effect on latewood width[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], which is more strongly controlled by summer drought stress[<!--@tardif_influence_2006;-->@roibu_climatic_2020;@kern_multiple_2013], suggesting that a positive effect of warm springs on the total C content of annual rings is unlikely. 
Rather, the most likely explanation of the observed lack of correlation between spring temperatures and stem growth is that growth is sink-limited, such that increased GPP does not enhance $ANPP_{woody}$[@delpierre_wood_2016].

<!--on the fate of extra C-->
Our results add to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@refs], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, fail to stimulate $ANPP_{woody}$[@jiang_fate_2020;@refs].
The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, $ANPP_{woody}$ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018;@buermann_widespread_2018].
Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be short lived.

@@

<!--the dominant effect of temperature on ANPP_woody is a negative effect of gs temperature-->
Regardless of the influence of spring temperatures on C cycling within the ecosystem, our results make clear that the dominant effect of temperatures on tree growth occur not in the spring, but during the peak growing season, when increased atmospheric demand associated with high temperatures limits growth[@novick_increasing_2016;@delpierre_wood_2016;@etzold_number_2022;@zweifel_why_2021]. 
Indeed, the timing of peak growth in June and July (`r EDT_phenology_parameters`, `r EDF_growth_curves_all`) coincides with the timing of the greatest sensitivity of annual growth to $T_{max}$ (Fig. `r tree_rings`, `r EDT_tree_ring_summary`).
This finding is consistent with numerous tree-ring studies demonstrating strong sensitivity of growth to drought stress or high temperatures during the peak growing season[@dorangeville_drought_2018;@helcoski_growing_2019;@anderson-teixeira_joint_2021;@babst_twentieth_2019].
<!--*It is also worth noting that growth is often influenced by temperature of the preceding year(s) through its influence on production and storage of NSCs[@refs;@helcoski], and indeed we observe many significant effects of $T_{max}$ in the preceding growing season, albeit very few in the preceding spring (`r EDF_tree_rings_TMAX_extended`).*--><!--this sentence feels tangential, but discussion of these lag effects was requested by one reviewer.-->
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021;@lian_summer_2022].
While such an interaction was unlikely to have played a major role within the scope of our dendrometer band study, given relatively mesic conditions and lack of significant correlation between spring temperatures and summer drought stress (see Methods), there is potential for warm springs to have a net negative effect on growth through such mechanisms at other times or places[@gao_earlier_2022].
